---
title: "Deaths in the field"
output:
  pdf_document: default
  html_notebook: default
---
```{r Welcome}

# Code written in R Markdown by Marjorie Cantine to analyze data about deaths during geological fieldwork (last updated 8 September 2021)
# Code input is the database Excel file (FieldDeaths_DATE.xls); output is several ggplot-generated graphs, saved as PDFs. The input needs to be in the same directory as this code; that is where the graphs will be saved, too. 
# Code also generates a number of summary tables, which don't produce outputs but can be queried individually. These are listed in Chunk 5: Geologist Death Subsetting 
# To run this code, either press Knit to generate a single PDF with all figures, or Run > Restart R and Run all Chunks to generate individual PDFs 
# Please contact me with any comments or corrections at mcantine@gmail.com

```
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

#Load the relevant libraries 
library("grid")
library("gridExtra")
library("ggplot2")
library("gdata")
library("dplyr")
library('forcats')
library('tinytex')
library('here')
library(EnvStats)
library('RColorBrewer')
library(gtable)
library(cowplot)

# Colorblind palette with grey:
cbPalette <- c("#E69F00", "#56B4E9", "#F0E442", "#009E73","#0072B2", "#D55E00", "#CC79A7", "#999999", "burlywood4", '#222222')

#Set working directory
here()

```

```{r Import and subset database, echo = FALSE}

# Import database 
Db <- read.xls('FieldDeaths_20200907.xlsx',  skip = 0) # Import database 
Db <- Db[,1:20] #trim to just active columns

# give columns short names 
names(Db) <- c("number", "name", "gender", "field", "stage.or.age", "sector", "short.cause", "detailed.cause", "category", "BLS.category", "year", "country.death", "country.worked", "earth.science", "in.the.field", "rationale", "perma.cc.link", "media.category", "source.name", "student") 


# Subset the database so that it only includes deaths that occurred during doing Earth science (earth.science == "Yes") and deaths that occurred "in the field" and not at a mine or construction site (in.the.field == "Yes")
Db.clean <- subset(Db, Db$earth.science == "Yes" & Db$in.the.field == "Yes")

#Ok. Database is ready to be analyzed 
    
```
```{r Fig1 Pie charts, echo = FALSE}


# Goal: Make pie chart of intentional vs. unintentional deaths.

#Define colors for use in pie chart 
colors <- c("intentional" = 'lightsteelblue4', "unintentional" = 'lightsteelblue1', "medical" = 'lightsteelblue2')

#Use dplyr functions to summarize number of deaths by cause and calculate proportions
death.category <- Db.clean %>% 
  group_by(category) %>% 
  count()   %>%
  mutate(prop = n / sum(death.category$n) *100)  

#add data source (this will be useful once we start comparing this data to other datasets)
death.intentional$source = c("this.database","this.database","this.database") 

#give columns short descriptive names
names(death.intentional) <- c("cause", "n", "prop",  "source")

#make pie chart
Pie_intentional <- ggplot(death.intentional, aes(x="", y=prop, fill=cause)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(text=element_text(size=8),
        panel.background = element_blank()) +
  ggtitle('Intentional death')

plot(Pie_intentional)


## Make a pie chart of deaths by gender. 

#define colors 
colors <- c("M" = 'darkseagreen1', "X" = 'darkseagreen3', 'F' = 'darkseagreen4')

#Use dplyr functions to get to a summary table that separates deaths by gender, calculate proportion
death.gender <- Db.clean %>% 
  group_by(gender) %>% 
  count() %>%
  arrange(desc(gender)) %>%
  mutate(prop = n / sum(death.gender$n) *100) 

#add source information
death.gender$source = c("this.database","this.database","this.database")

#make pie chart
Pie_gender <- ggplot(death.gender, aes(x="", y=prop, fill=gender)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(
        text=element_text(size=8),
        panel.background = element_blank())+
  ggtitle('Gender and death')

plot(Pie_gender)


## Make a pie chart of deaths by sector 

#First, let's clean up the sector column and group everything except for academia, private, and government into "other"
Db.clean$sector.coded <- Db.clean$sector
Db.clean$sector.coded[Db.clean$sector.coded == "Personal"] <- "Other"
Db.clean$sector.coded[Db.clean$sector.coded == "Unknown"] <- "Other"
Db.clean$sector.coded[Db.clean$sector.coded == "School"] <- "Other"
Db.clean$sector.coded[Db.clean$sector.coded == "Non-profit"] <- "Other"

#define colors for plotting
colors <- c("Academia" = 'gray10', "Private" = 'gray32', 'Government' = 'gray73', "Other" = "gray93")

#dpylr functions to get to a clean summary table 
death.sector <- Db.clean %>% 
  group_by(sector.coded) %>% 
  count() %>%
  arrange(desc(sector.coded)) %>%
  mutate(prop = n / sum(death.sector$n) *100) 

#make pie chart
Pie_sector <- ggplot(death.sector, aes(x="", y=prop, fill=sector.coded)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(text=element_text(size=8),
        panel.background = element_blank()) +
  ggtitle('Sector and death')

plot(Pie_sector)


#arrange these three plots into one unit and output it as a PDF 
Geologist_pies_plot <- grid.arrange(
  grobs = list(Pie_intentional, Pie_gender, Pie_sector),
  widths = c(1),
  heights = c(1,1,1),
  layout_matrix = rbind(c(1),
                        c(2),
                        c(3)))

ggsave("Geologist_pies_plot.pdf", Geologist_pies_plot) 

```

```{r Geologist death subsetting}

#These summary tables don't output charts, but they calculate various proportions. 

# look at intentional death death by sector and gender 
Db.intentional <- subset(Db.clean, category == "intentional")

death.intentional.sector <- Db.intentional %>% 
  group_by(sector.coded) %>% 
  count()

death.intentional.gender <- Db.intentional %>% 
  group_by(gender) %>% 
  count()

# Look at gender breakdown of deaths by sector 

Db.clean.academia <- subset(Db.clean, sector.coded == "Academia")
Db.clean.private <- subset(Db.clean, sector.coded == "Private")
Db.clean.government <- subset(Db.clean, sector.coded == "Government")

death.academia.gender <- Db.clean.academia %>% 
  group_by(gender) %>% 
  count()

death.private.gender <- Db.clean.private %>% 
  group_by(gender) %>% 
  count()

death.government.gender <- Db.clean.government %>% 
  group_by(gender) %>% 
  count()

```

```{r Fig1 Bar chart}

#Goal is to make a figure that summarizes the causes of unintentional death

# Find all unintentional deaths 
Db.unintentional <- subset(Db.clean, intentional == "unintentional")

#summary table of causes
death.unintentional <- Db.unintentional %>% 
  group_by(short.cause) %>% 
  count()  

#refactor the data for plotting purposes 
death.unintentional$short.cause <- factor(death.unintentional$short.cause, 
                                          levels = c("vehicle", "helicopter", "animal", "drowning", "fall", "volcanic hazard", "heat", "SCUBA", "avalanche", "electrocution", "rock fall", "heavy machinery" ))

#make the plot 
Unintentional_plot <- ggplot(data=death.unintentional, aes(x=short.cause, y=n)) +
  geom_bar(stat="identity", fill="darkslateblue") +
  geom_text(aes(label=n), vjust=-0.3, size=3.5) +
  theme_bw()+
  theme(aspect.ratio= 1/3,
        text = element_text(family = "Helvetica", color = "black"),
        panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 15)) +
  xlab("Cause of death") +
  ylab("Number")

plot(Unintentional_plot)
ggsave("Unintentional_plot.pdf", Unintentional_plot, width = 5, units = "in")


```
```{r Fig2 Pie charts, echo = FALSE}

#Goal in this section is to plot data from the NPS and BLS (causes and sex/gender), and to add another chart looking at geologist deaths using the BLS categories 

# National Parks data 
# From https://perma.cc/U3XX-G5GD 
# and from https://perma.cc/UN5H-35HH -- CY2014-CY2016 Mortality Dashboard Key Statistics
# Reported: 79% of deaths are male 
# 53% (525) of deaths are unintentional; 22% (222) are medical; 18% (174) intentional (and of those 95% suicide); 7% (69) are undetermined 

#make dataframe about sex in NPS data 
nps.data.sex <- tibble(c("male", "female", "undetermined"), c(79, 19, 2))
names(nps.data.sex) <- c("sex", "fraction")

#make dataframe about cause in NPS data 
nps.data.cause <- tibble(c("unintentional", "medical", "intentional", "undetermined"), c(53, 22, 18, 7), c(525, 222, 174, 69), c("nps","nps","nps","nps"))
names(nps.data.cause) <- c("cause", "fraction", "n", "source")


# BLS data 
# From https://perma.cc/K9LZ-QTC7 -- Fatal occupational injuries by worker characteristics and event or exposure, 2019 (latest available)
# Reported: 5,333 total deaths
# Reported gender breakdown: 437 women; 4896 men, adds up to 5333
# Reported causes: 2122 transportation; 841 violence and injuries from humans and animals; 732 contact with objects and equipment; 880 falls, trips, and spills; 642 exposure to harmful substances or environments; 99 fires and explosions. This does not add up to 5333--17 deaths (less than 1%) not categorized

#make dataframe about gender in BLS data 
bls.data.gender <- tibble(c("man", "woman"), c( 4896/5333 * 100, 437/5333 * 100), c(4896, 437), c('bls','bls'))
names(bls.data.gender) <- c("gender", "fraction", 'n', 'source')

#make dataframe about cause in BLS data 
bls.data.cause.source <- c('bls','bls','bls','bls','bls','bls')
bls.data.cause.n <- c(732, 880, 99, 642, 2122, 841)
bls.data.cause.percent <- c(732/5316 * 100, 
                      880/5316 * 100, 
                      99/5316 * 100,
                      642/5316 * 100,
                      2122/5316 * 100, 
                      841/5316 * 100)
bls.data.cause <- tibble(c("contact",
                           "falls", 
                           "explosions",
                           "exposure",
                           "transportation", 
                           "violence"), 
                         bls.data.cause.percent, bls.data.cause.n, bls.data.cause.source)
names(bls.data.cause) <- c("BLS.category", "fraction", "n", "source")

## Make a pie chart of deaths by gender. 
#NPS data first 

#define colors
colors <- c("male" = 'darkseagreen1', "undetermined" = 'darkseagreen3', 'female' = 'darkseagreen4')

#make pie chart
NPS_sex <- ggplot(nps.data.sex, aes(x="", y=fraction, fill=sex)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(text=element_text(size=8),
        panel.background = element_blank()) +
  ggtitle('Sex and death, NPS data')

plot(NPS_sex)

# BLS data next
#define colors
colors <- c("man" = 'darkseagreen1', 'woman' = 'darkseagreen4')

#make pie chart
BLS_gender <- ggplot(bls.data.gender, aes(x="", y=fraction, fill=gender)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(text=element_text(size=8),
        panel.background = element_blank()) +
  ggtitle('Gender and death, BLS data')

plot(BLS_gender)

# Now plot causes 

# NPS first 
#define colors 
colors <- c("intentional" = 'lightsteelblue4', "unintentional" = 'lightsteelblue1', "medical" = 'lightsteelblue2', "undetermined" = "gray50")

#make pie chart
NPS_cause <- ggplot(nps.data.cause, aes(x="", y=fraction, fill=cause)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(text=element_text(size=8),
        panel.background = element_blank()) +
  ggtitle('NPS death by cause')

plot(NPS_cause)

# Now BLS causes 
#define colors 
colors <- c("transportation" = 'gray90', 
            'violence' = 'gray70',
            'contact' = 'gray5',
            'exposure' = 'gray40',
            'falls' = 'gray60',
            'explosions' ='gray15')

#make pie chart
BLS_cause <- ggplot(bls.data.cause, aes(x="", y=fraction, fill=BLS.category)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(text=element_text(size=8),
        panel.background = element_blank()) +
  ggtitle('BLS data by cause')

plot(BLS_cause)


# Now do the geology data with BLS categories 

#subset data for those deaths that could be categorized using BLS categories
death.bls.analysis <- subset(Db.clean, BLS.category != "NA")

#define colors
colors <- c("transportation" = 'gray90', 
            'violence' = 'gray70',
            'contact' = 'gray5',
            'exposure' = 'gray40',
            'falls' = 'gray60',
            'explosions' ='gray15')

#generate summary table
death.bls.analysis <- death.bls.analysis  %>% 
  group_by(BLS.category) %>% 
  count()

#calculate proportion
death.bls.analysis  <- death.bls.analysis  %>%
  arrange(desc(n)) %>%
  mutate(prop = n / sum(death.bls.analysis$n) *100) 

#add note about source for later use 
death.bls.analysis$source = c('this.database','this.database','this.database','this.database','this.database')

#make pie chart
BLS_geology_pie <- ggplot(death.bls.analysis, aes(x="", y=prop, fill=BLS.category)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_manual(values=colors) +
  coord_polar("y", start=0)+
  theme_void()+
  theme(text=element_text(size=8),
        panel.background = element_blank()) +
  ggtitle('Geologist data, BLS categories')

plot(BLS_geology_pie)


ggsave("NPS_sex.pdf", NPS_sex, width = 2, height = 2, units = "in") 
ggsave("NPS_cause.pdf", NPS_cause, width = 2, height = 2, units = "in") 
ggsave("BLS_gender.pdf", BLS_gender, width = 2, height = 2, units = "in") 
ggsave("BLS_cause.pdf", BLS_cause, width = 2, height = 2, units = "in") 
ggsave("BLS_geology_pie.pdf", BLS_geology_pie, width = 2, height = 2, units = "in") 


```
```{r ANOVA test}

#Compile BLS and geologist data together
bls.cause.comp <- full_join(bls.data.cause, death.bls.analysis) 

#Run ANOVA
bls.cause.results <- aov(n ~ BLS.category + source, data = bls.cause.comp)
summary(bls.cause.results)

#Compile NPS and geologist data
nps.cause.comp <- full_join(nps.data.cause, death.intentional) 

#Run ANOVA
nps.cause.results <- aov(n ~ cause + source, data = nps.cause.comp)
summary(nps.cause.results)
